A Case Study in Preserving a High Energy Physics Application

### 2014-03-14 paper.tex
Rewrote section 3.

\documentclass{acm_proc_article-sp}
\usepackage{url}
\usepackage[flushleft]{threeparttable}
\begin{document}

@@

The largest of these repositories is the CMS Software Distribution (CMSSW),
a carefully-curated selection of software packages which is distributed
in binary form.  Historically, CMSSW was downloaded and installed on shared
filesystems within HPC centers.  In recent years, distribution has moved to
an on-demand delivery system known as CVMFS~\cite{blomer2011cernvm}.  The content
of CMSSW is managed very carefully by a centralized team whose main goal
is to ensure that the current version of the software operates correctly
on the operating systems and architectures currently in use.  However,
preservation is not an objective of the group, and so there is
no guarantee that old versions of CMSSW operate in new environments,
or vice versa.

\emph{TauRoast} was provided to us in the form of an
email which described, in prose, how to obtain the source,
build the program, and run it correctly on one specific
machine at Notre Dame, with no particularly guarantee that
it will run anywhere else in the world.
Although this starting point may seem extreme, it is
perfectly natural for collaborators to share configurations
with each other in this form, and to rely on the presence
of a working environment with appropriate dependencies already
installed.

The authors played the role of curators, whose job it is to prepare the application for permanent archival by determining and packaging the dependencies.

\subsection{Workflow of the Example}

The workflow of the exmaple can be divided into eight procedures.

(a) Declare environment variables

(b) Obtain software from CMSSW

(c) Obtain software from Git

(d) Obtain software from some public HTTP web links

(e) Obtain software from one private home page

(f) Build software enviroment using SCRAM ~\cite{rabbertz2004cms} and Python

(g) Install grid control ~\cite{holtman2001cms}, obtain CMS data from Grid and store the data into HDFS ~\cite{borthakur2008hdfs} mounted as one local file system.

(h) Actual data analysis.

\subsection{Key Observations of the Example}

\subsubsection{Complexity of Data and Software Dependencies}

\begin{figure}[t]
\centering
\includegraphics[width=.8\columnwidth]{dependencies.eps}
\caption{Dependencies}
\label{fig:dependencies}
\end{figure}

Figure~\ref{fig:dependencies} illustrates all the dependencies
involved in the program, which can be divided into two categories: clear
dependencies and implicit dependencies. Clear dependencies include scientific data
dependencies and direct software dependencies which can be extracted directly
from the analysis program. The scientific data of this example comes from the Grid ~\cite{holtman2001cms}. Direct software dependencies include CMSSW, Git, HTTP and one personal home page.

Implicit dependencies include indirect software
dependencies which is necessary for the successful execution of direct software
dependencies and other implicit dependencies. For example, we know Git is one
direct software dependency from the analysis program, which further depends on
perl, python, rsync, openssh-clients and other packages. The analysis program
accesses one file called File1, which is one symbolic link file further referring to
another file named File2.
The implicit dependencies also include some remote file systems mounted as local file systems, such CVMFS, AFS ~\cite{howard1988scale}, NFS ~\cite{sandberg1985design}, PanFS ~\cite{welch2008scalable} and HDFS.
More importantly, the successful execution of this program greatly depends on the underlying OS and hardware platform.

\subsubsection{Size of Data and Software Dependencies}
The size of data and software dependencies is astonishing.
The size of BEAN is 11.6TB, which is too large to be analyzed directly by the tau roast program. After one preprocessing which only collects the events matching basic quality criteria, the data size is still very large, 43.3GB.

The first column of Table~\ref{table:size-original-real} illustrates the size of each source of data and software. The size of PanFS is as large as 155TB and the size of Ntuples data is 24TB. The total size of all the dependencies is about 181TB, which directly makes the idea of preserving the original execution environment completely impossible.

\subsubsection{Complexity of the Program}
Complex data and software dependencies, complex environment building process, the large size of data and software involved in the tau roast program, together with the access authority problem, determine the complexity of the program. 
In order to obtain the experimental data from the Grid, the author needs to install and config grid-control software. 
The access authority of the Grid is also necessary for the data acquisition.
The software acquisition from CMSSW, Git and HTTP and environment building process is time-consuming. The execution time of actual data analysis is about 20 minutes. However, the software acquisition and environment building time, excluding data acquisition time, is about 14 minutes. If the acquisition of experimental data is considered, actual data analysis only occupies one small percentage of the whole time consumption.
The size of data and software dependencies further complicates the program.

\subsubsection{Really Used Data and Software Size}

Even if the orginal data size referred by the program is astonishing, the size
of really used data and software is greatly smaller.
Figure~\ref{fig:data-code-size} illustrates the inputs of the tau roast program.
After the TauAnalysis procedure, the data size of Ntuples is 43.3GB, however, the atual size of really used
data in the program is 20GB.  The same trend occurs to the size of software and
other dependencies. The size of CMSSW repository is 13.1GB, the size of
packages checked out from CMSSW is 448.3MB, and the size of actually used CMSSW
files is only 6.3MB. 
The total size of PanFS mounted as {\tt /pscratch} is 155TB, however, the really used data from {\tt /pscratch} is only 1.6MB.
The third column of Table~\ref{table:size-original-real} illustrates the size of actually used data and software from each dependency.

\subsubsection{Stability of Dependencies}

The program involves obtaining software from the home page of another graduate
student from University of Notre Dame. However, the maintenance of this
home page will terminate after the graduation of the student.
The stability of each dependency must be evaluated to ensure the program can be 
repeated by others.

The stability of the Grid is relatively higher. However, the
public web resources and personal home page is not stable. Too many factors
like the termination of one web site, the graduation of the home page's author,
would result in the failure of resource access. CMSSW, which seems to be
stable, in fact, also involves stability problems. CMSSW provides different
collections of accessible versions to different architectures. Transformation
from one architecture to another may result in the failure of accessing one
certain CMSSW version. At the beginning of our reseach, we can use the cvs command to
check out packages from CMSSW anonymously. However, this anonymous access has
been cancelled now.

@@

\begin{figure*}
\centering
\includegraphics[width=1.6\columnwidth]{version-evolution.eps}
\caption{Version Evolution}
\label{fig:version-evolution}
\end{figure*}

>>>>

\documentclass{sig-alternate}
\usepackage{url}
\usepackage[flushleft]{threeparttable}
\begin{document}

@@

The largest of these repositories is the CMS Software Distribution (CMSSW),
a carefully-curated selection of software packages which is distributed
in several forms.  Historically, components of CMSSW were obtained by checking components
of the source out of CVS, or by installing a complete binary package on a shared
filesystem within an HPC center.  In recent years, distribution has moved to
an on-demand delivery system known as CVMFS~\cite{blomer2011cernvm}.  The content
of CMSSW is managed very carefully by a centralized team whose main goal
is to ensure that the current version of the software operates correctly
on the operating systems and architectures currently in use.  However,
preservation is not an objective of the group, and so there is
no guarantee that old versions of CMSSW operate in new environments,
or vice versa.

\section{Observations}

\emph{TauRoast} was provided to us in the form of an
email which described, in prose, how to obtain the source,
build the program, and run it correctly on one specific
machine at our home institution, with no particular guarantee that
it will run anywhere else in the world.
Although this starting point may seem extreme, it is
perfectly natural for collaborators to share configurations
with each other in this form, and to rely on the presence
of a working environment with appropriate dependencies already
installed.  From this starting point, the authors played the
role of curators, whose job it is to prepare the application
for permanent archival.

First, we elaborated the email instructions into an
executable script that obtains the dependencies and then
executes the analysis.  The script declares the necessary
environment variables, downloads and checks out the necessary source code,
builds it appropriately, calls initialization scripts in
the dependent software, and then runs the analysis.
A few rounds of correction with the original author were necessary
to obtain all the dependencies and run the artifact correctly.
(The original email also indicated how to run the application
within a production batch system.  For the purposes of preservation,
we consider the execution infrastructure to be distinct from the application,
and leave it out of consideration for now.)

The process of elaborating the program into a script revealed
several observations about this type of application:

\begin{itemize}

\item {\bf Many Explicit External Dependencies.}  TauRoast depends on a large number of
external dependencies, each with a different access method and data source.
While we knew in advance that it depended upon the large CMSSW distribution,
it was not apparent until elaborating the script that it depended upon
two different Github repositories for the Tau source,
a CVS server at CERN for some configuration information, a public web page
for the PyYAML library, and the public home page of a Notre Dame student
for one missing header file.  (The latter is particularly troubling!)
While, at some level, the authors and users of these software know of these dependencies, they are often missing in
informal communications or forgotten once the dependency is install.
However, once known, they are at least expressed explicitly within the script.

{\bf Many Implicit Local Dependencies.} A much harder problem is that the
application assumed the presence of many different components in the local
filesystem view.  It would be tempting to capture all of these by simply
storing a virtual machine image containing the local filesystem.  However,
the application dependend on no less than {\bf six} networked filesystems
available on a particular machine available to the author:
the data to be analyzed was stored on an HDFS~\cite{hadoop} cluster,
some configuration data was stored on a CVMFS~\cite{cvmfs} filesystem,
and a variety of software tools were on an NFS~\cite{howard1988scale},
pNFS~\cite{welch2008scalable} and AFS~\cite{sandberg1985design} systems.
The original authors were not aware of many of these dependencies,
because they simply relied on local administrators to configure the
software and make it available.

\item {\bf Configuration Complexity.}  As a means of controlling the complexity
of dependent software packages, the high energy physics community has developed
a number of tools that perform run-time configuration and consistency checks
of the available software.  {\tt scram} is the software management tool used
by the CMS experiment.  Before running any code, {\tt scram} is used to locate
the appropriate version software,  set environment variables such as the PATH, run any
tool-specific configuration, and do the same for all software on which it depends.
If the correct versions are not available, {\tt scram} halts and emits an error.
While this procedure has great value for consistency, it also introduces a significant cost
because it involves a large number of nested scripts traversing a filesystem,
repeatedly looking up metadata.  In our example, the time to perform this configuration
with a cold cache is about 14 minutes, which is almost as large as the actual analysis
run, which takes 20 minutes.

{\bf High Selectivity.}  Although the total size of the resources accessed by this
program is very large, the size of the data and software actually used are much smaller.
Often, an entire repository or data source is named within the script, but the program
only needs a handful of items from that source.  For example, the data is stored on an
HDFS filesystem with 24TB of data, but only 43GB of that are the input Ntuples,
and of those only 20GB are actually consumed by the program. The CMSSW repository is 88GB in total
but only 448MB in source are checked out, and the software actually used only
measured 6.3MB.  In a few cases, a source of software is named but never actually accessed.
(For example, our original script includes the Open Science Grid software stack in the PATH, but does not actually use it.)
We suspect that end users are accustomed to missing dependencies and thus get in the habit of adding commonly used software,
whether it is needed or not.

\item {\bf Rapid Changes in Dependencies.}  Over the course of three months
between collecting the initial email, analyzing the program, and writing this
paper, the computing environment was under continuous change.  The CMSSW software
distribution released a new version, the target execution environment was upgraded
to a new operating system, and the experiment deprecated the use of CVS for obtaining
the software.  While the users of this software seem be accustomed to constant change,
any preservation technique will have to be very cautious about relying upon an
external service, even one that may appear to be highly stable.

\end{itemize}

\if 0

@@

\begin{figure*}
\centering
\includegraphics[width=1.6\columnwidth]{version-evolution.eps}
\caption{Version Evolution}
\label{fig:version-evolution}
\end{figure*}
\fi
